To Batch or Not to Batch: Sequential vs. Batched Testing Strategies in the ED

### 2023-11-07 manuscript/manuscript.tex
mediation analysis

Our analytical lens is focused on the Emergency Department at the Mayo
Clinic of Arizona, a distinguished tertiary care establishment. During
our study's timeframe, the ED recorded an annual visitation of
approximately 41,000 patients. The department is singularly staffed by
board-eligible or board-certified emergency physicians, abstaining from
the services of nurse practitioners or physician assistants. A notable
observation was that residents in rotation oversaw a low fraction,
roughly 10\%, of the patient volume. Comprehensive patient data,
encompassing demographics, chief complaints, vital signs, emergency
severity, length of stay, and resource utilization metrics, were
meticulously logged during the study period.

@@

In the context of diagnostic test ordering, we defined batching as
placing multiple test orders within 5 minutes of one another.
Sensitivity analyses on this cutoff point showed that our results are
robust to this definition. Our study differentiated between two distinct
batching strategies:

\begin{itemize}
\item Lab + Image Batch: The batch order for a patient consisted of two or more distinct diagnostic tests where at least one imaging (contrasted CT scan, non-contrasted CT scan, X-ray, ultrasound) and one lab test were placed.

\item Image + Image Batch: The batch order for a patient consisted of two or more distinct imaging (contrasted CT scan, non-contrasted CT scan, X-ray, ultrasound) diagnostic tests.
\end{itemize}

The distinction is necessary because laboratory tests may be processed
in parallel with imaging tests, which have a much more limited effect on
patient waiting time. However, a batch of multiple imaging tests
requires that the patient is present for both tests. Thus, waiting times
are much more likely to be impacted. Therefore, the specific tests that
make up a batch may provide insight into how the physician trades off
between waiting times and uncertainty, given the various complexity
levels of patient encounters. Figure 1 shows the rate at which each
batching type occurs for the top ten most frequently occurring chief
complaint fields.

\begin{figure}[h]
  \centering
  \includegraphics[width=1\textwidth]{../manuscript/figures/frequency_batches.png}
  \tiny
  \caption{Figure 1 displays two statistics for the ten most common major chief complaint categories observed in our ED visit sample: the unadjusted batching rate for lab-image batching and image-image batching.}
\end{figure}

\hypertarget{statistics}{%
\subsection{Statistics}\label{statistics}}

Table 2 provides a comprehensive breakdown of the patient encounters
involving diagnostic tests ordered individually in a batched or
sequential way during an ED encounter from our baseline sample. About
39\% of patient encounters involved diagnostic tests ordered as part of
a batch, most of which (64\%) included two tests.

Notable distinctions emerge when comparing the two cohorts. Patients
subjected to batched test ordering exhibited a notably prolonged stay in
the emergency department (ED) (p\textless0.001) and were more frequently
assigned a lower Emergency Severity Index (ESI). Furthermore, this group
skewed towards older demographics, with half of them falling into the
65+ age category. Patients in the batch-order group also displayed a
higher likelihood of presenting symptoms at triage, including
tachycardia (p\textless0.001), tachypnea (p\textless0.001), fever
(p\textless0.001), and hypotension (p\textless0.001). Delving deeper
into the ESI data unveils a higher concentration of patients with more
critical conditions (ESI 1 and 2) within the batched group, in contrast
to the sequential group (45.1\% with ESI of 1 or 2 in the batched group
compared to 27.5\% with ESI of 1 or 2 in the sequential group). This
finding suggests that factors related to patient complexity and severity
(Saghafian et al., 2014) significantly influence physicians' decisions
to opt for batch-ordering of diagnostic tests. However, it is important
to note that our investigation revealed substantial variation among
physicians in their choices between batching and sequencing for a given
patient.

\begin{table}
\centering
\caption{Summary Statistics}
\begin{tabular}{p{3cm} *{4}{p{2.5cm}}p{2.5cm}p{2.5cm}p{2.5cm}p{2.5cm}p{2.5cm}p{2.5cm}p{2.5cm}p{2.5cm}}
\toprule
Variable & Overall Visits & Sequentially Arm & Batch Arm & $p$-value \\
&  ($N = 41,846$) & ($N = 25,354$)  & ($N = 16,492$) & \\
\midrule
ESI & & & & $<0.001$ \\
1 & 476 (1.1\%) & 136 (0.5\%) & 340 (2.1\%) & \\
2 & 14,020 (34\%) & 6,865 (27\%) & 7,155 (43\%) & \\
3 & 24,006 (57\%) & 15,517 (61\%) & 8,489 (52\%) & \\
4 & 3,283 (7.9\%) & 2,787 (11\%) & 496 (3.0\%) & \\
5 & 34 (<0.1\%) & 33 (0.1\%) & 1 (<0.1\%) & \\
(Missing) & 27 & 16 & 11 & \\
\midrule
Age & & & & $<0.001$ \\
<20 & 1,413 (3.4\%) & 1,154 (4.6\%) & 259 (1.6\%) & \\
20-45 & 9,419 (23\%) & 6,538 (26\%) & 2,881 (17\%) & \\
45-65 & 12,829 (31\%) & 7,764 (31\%) & 5,065 (31\%) & \\
65+ & 18,185 (43\%) & 9,898 (39\%) & 8,287 (50\%) & \\
\midrule
ED LOS (min) & & & & $<0.001$ \\
Mean (SD) & 269 (379) & 255 (174) & 290 (564) & \\
\midrule
Disposition & & & & $<0.001$ \\
Admit & 9,124 (22\%) & 4,286 (17\%) & 4,838 (29\%) & \\
Discharge & 25,861 (62\%) & 17,596 (69\%) & 8,265 (50\%) & \\
Other & 6,861 (16\%) & 3,472 (14\%) & 3,389 (21\%) & \\
\midrule
Tachycardic & 8,328 (20\%) & 4,853 (19\%) & 3,475 (21\%) & $<0.001$ \\
Tachypneic & 3,953 (9.4\%) & 1,818 (7.2\%) & 2,135 (13\%) & $<0.001$ \\
Febrile & 1,013 (2.4\%) & 386 (1.5\%) & 627 (3.8\%) & $<0.001$ \\
Hypotensive & 69 (1.6\%) & 327 (1.3\%) & 342 (2.1\%) & $<0.001$ \\
\midrule
Number of Tests & & & & $<0.001$ \\
1 & 13,406 (32\%) & 13,406 (53\%) & 0 (0\%) & \\
2 & 19,291 (46\%) & 8,782 (35\%) & 10,509 (64\%) & \\
3 & 7,707 (18\%) & 2,689 (11\%) & 5,018 (30\%) & \\
4 & 1,348 (3.2\%) & 451 (1.8\%) & 897 (5.4\%) & \\
5 & 94 (0.2\%) & 26 (0.1\%) & 68 (0.4\%) & \\
\midrule
Gender & & & & $<0.001$ \\
Female & 22,441 (54\%) & 14,053 (55\%) & 8,388 (51\%) & \\
Male & 19,405 (46\%) & 11,301 (45\%) & 8,104 (49\%) & \\
\midrule
Race & & & & $0.002$ \\
White & 37,070 (89\%) & 22,355 (88\%) & 14,715 (89\%) & \\
Black & 1,727 (4.1\%) & 1,057 (4.2\%) & 670 (4.1\%) & \\
Asian & 1,260 (3.0\%) & 775 (3.1\%) & 485 (2.9\%) & \\
Native & 546 (1.3\%) & 354 (1.4\%) & 192 (1.2\%) & \\
Other & 763 (1.8\%) & 493 (1.9\%) & 270 (1.6\%) & \\
Unknown & 480 (1.1\%) & 320 (1.3\%) & 160 (1.0\%) & \\
\bottomrule
\multicolumn{5}{l}{\textsuperscript{1} n (\%)} \\
\multicolumn{5}{l}{\textsuperscript{2} Pearson's Chi-squared test; Welch Two Sample t-test} \\
\end{tabular}
\end{table}

@@

Contrary to most healthcare settings where patients exhibit choice, they
are predominantly passive in their physician assignment in the ED. In
most EDs, however, physicians have discretion in picking their patients.
In contrast, patients arriving at the Mayo Clinic ED are randomly
assigned to physicians via a rotational patient assignment algorithm
(Traub et al., 2016), which removes potential selection bias concerns
for our analyses. In essence, barring arrival time and shift-level
variation, the physician-to-patient matching can be deemed random. Table
2 displays that patient encounters (regarding chief complaints and
emergency severity) are equitably distributed across physicians within
our study's cohort.

\begin{table}[htbp]
    \centering
    \caption{Wald Test Results}
    \begin{tabular}{lcc}
        \toprule
        Chief Complaints & F-Statistic & $Pr(> F)$ \\
        \midrule
        Abdominal Complaints & 1.37 & 0.106 \\
        Back or Flank Pain & 1.00 & 0.451 \\
        Chest Pain & 0.98 & 0.476 \\
        Extremity Complaints & 0.97 & 0.495 \\
        Falls, Motor Vehicle Crashes, Assaults, and Trauma & 0.73 & 0.812 \\
        Gastrointestinal Issues & 0.98 & 0.480 \\
        Neurological Issue & 0.75 & 0.793 \\
        Shortness of Breath & 1.23 & 0.199 \\
        Skin Complaints & 1.05 & 0.388 \\
        Upper Respiratory Symptoms & 1.21 & 0.218 \\
        \midrule
        Emergency Severity & F-Statistic & $Pr(> F)$ \\
        \midrule
        ESI 1 or 2 & 1.09 & 0.346 \\
        ESI 3, 4, or 5 & 1.24 & 0.196 \\
        \bottomrule
    \end{tabular}
\begin{tablenotes}
\tiny
\item Table 2 reports the results of a Wald test which was conducted to assess the balance of chief complaints across providers in our dataset. A balanced distribution implies that complaints and severity are evenly distributed across providers, which we expect to be the case due to randomization. The Wald F-statistic and p-value are reported. Robust standard errors (type HC1) were used to account for potential heteroscedasticity in the data.
\end{tablenotes}
\end{table}

Our study employs the ``judges design'', a quasi-random assignment
strategy widely used in literature to investigate causal effects.
Typically, this design exploits variations in sentencing leniency among
judges working in the same court. We adopt a similar approach by
examining batching variations across physicians working in the same
emergency department. This method enables us to identify the causal
impact of being assigned to different types of physicians with differing
batch-ordering tendencies.

@@

\begin{equation}
Batched_{i,t} = \alpha_0 + \alpha_{ym} + \alpha_{dt} + \varepsilon_{i,t}
\end{equation}

Where \(Batched_{i,t}\) is a dummy variable equal to one if patient
\(i\) had their diagnostic tests batched on encounter that took place on
data \(t\). Fixed effects include year-month fixed effects,
\(\alpha_{ym}\), to control for time and seasonal variation in batching,
such as hospital-specific policies (e.g.~initiatives to eliminate excess
testing) or seasonality in ED visits. We also control for
``shift-level'' variations that include both physician scheduling and
patient arrival with day of week-time of day fixed effects,
\(\alpha_{dt}\). As stated earlier, these controls are what is required
for our quasi-random assignment assumption. Under the assumption that we
have captured the observables under which quasi-random assignment occurs
in the ED, the unexplained variation-- the physician's contribution--
resides in the error term, \(\varepsilon_{i,t}\).

@@

\begin{equation}
Tendency_{i,j}^{phys} =
\frac{1}{N_{-i,j}} \sum_{i' \in \{J \backslash i\}}\hat{e}_{i',k}
\end{equation}

where \(\hat{e}_{i',k} = \hat{Batch}_{i',k} - Batch_{i',k}\) is the
residual from equation (1); \(J\) is the set of all ED encounters
treated by physician \(j\); and \(N_{-i,j} = |\{J \backslash i\}|\), the
number of cases that physician has seen that year, excluding patient
\(i\). This leave-out mean eliminates the mechanical bias that stems
from patient \(i\)'s own case entering into the instrument. The measure
is interpreted as the average (leave-out) batch rate of patient \(i\)'s
physician, relative to other physicians in that hospital-year-month,
hospital-day of week-time of day.

We document that the Mayo Clinic ED physicians exhibit wide, systematic
variation in their propensity to batch order diagnostic tests. Figure 2
graphs the histogram of batch-ordering frequency by physician for
populat chief complaints, highlighting that the variation in batching
differs systematically. Table 3 presents the ``first stage'' in a
regression table: being assigned to a 10 pp higher tendency physician is
associated with a 10 pp increase in the likelihood of having tests
batch-ordered in the ED. The F-statistic is 51 when all controls and
fixed effects are included.

\begin{table}[htbp]
\centering
\caption{First Stage Regression Results}
\begin{threeparttable}
\begin{tabularx}{0.85\textwidth}{@{}Xccc@{}}
\toprule
 & Model 1 & Model 2 & Model 3 \\
\midrule
Batching Tendency & 0.99*** & 0.99*** & 0.99*** \\
 & (0.01) & (0.01) & (0.01) \\
\midrule
F statistic (full model)&7.32&52.03&50.93\\  
F (full model): p-value& $<0.001$ & $<0.001$ & $<0.001$ \\
\midrule
Num. obs. & 43300 & 43300 & 43300 \\
Seasonality and shift fixed effects? & Yes & Yes & Yes \\
Chief Complaint? & No & Yes & Yes \\
Patient observables? & No & No & Yes \\
\bottomrule
\end{tabularx}
\begin{tablenotes}
\tiny
\item Estimates of the first stage for the baseline sample described in the text. Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects. Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter. Patient observables include sex dummy, race/ethnicity, and age bins. Column 3 corresponds to the baseline controls. Robust standard errors are clustered at the physician level.
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}\tiny
\end{threeparttable}
\label{table:regression}
\end{table}

@@

>>>>

Our analytical lens is focused on the Emergency Department at the Mayo
Clinic of Arizona, a distinguished tertiary care establishment. During
our study's timeframe, the ED recorded an annual visitation of
approximately 45,000 patients. The department is singularly staffed by
board-eligible or board-certified emergency physicians, abstaining from
the services of nurse practitioners or physician assistants. A notable
observation was that residents in rotation oversaw a low fraction,
roughly 10\%, of the patient volume. Comprehensive patient data,
encompassing demographics, chief complaints, vital signs, emergency
severity, length of stay, and resource utilization metrics, were
meticulously logged during the study period.

@@

Our research design focuses on adults who visit the Mayo Clinic of
Arizona ED. We observe approximately 45,000 such visits until during the
stufy period. To improve power, we drop encounters with rare chief
complaints (\textless{} 1000 total encounters of this kind) and
complaints where a batch order occurs less than 10 percent of the time.
Since batch orders are rare for these cases, our physician batch
tendency instrument could suffer from a weak instrument problem were we
to include them. Example complaints dropped include urinary complaints
and \_\_\_\_\_. Excluding these conditions does not introduce selection
bias only if physician test batching tendency is orthogonal to physician
diagnosing behavior. While this assumption may be violated if we were to
use a very detailed level of chief complaint information upon which to
base our exclusion criterion, it is plausibly satisfied when using broad
complaint
categories\footnote{The idea being that, for a patient who presents with an ankle fracture, different physicians may choose different detailed diagnosis codes within the three-digit ICD-9 824 (“Fracture of ankle”), but are unlikely to disagree at the broader chief complaint cateogry.}.

\hypertarget{variable-definitions}{%
\subsection{Variable Definitions}\label{variable-definitions}}

\hypertarget{outcomes}{%
\subsubsection{Outcomes}\label{outcomes}}

In our instrumental variables (IV) analysis, the key explanatory
variable, \(Tendency_i\), is an indicator for whether patient \(i\) has
their tests batch ordered at their ED encounter. This variable is
crucial for understanding the causal effects of batch testing on various
health outcomes and resource utilization. Below, we detail our main
outcomes: (a) ED length of stay (LOS), (b) resource utilization, and (c)
72-hour return.

\emph{ED length of stay (LOS).}-- ED-LOS is a critical measure of
efficiency and patient throughput in emergency care settings. It is
defined as the duration from a patient's arrival to the ED until their
departure, whether by discharge or admission to the hospital. Although
literature provides a range for long ED-LOS, for the purpose of this
study, we operationalize long ED-LOS as stays exceeding the 75th
percentile in our dataset. This approach allows us to contextualize the
definition of long stays according to the specific demographics and case
mix of the EDs under study.

\emph{Resource Utilization}-- Resource utilization in the ED context
typically refers to the extent of medical services and interventions a
patient receives. In this study, we quantify resource utilization by the
total number of diagnostic tests ordered per patient during their ED
stay. This encompasses both initial and any subsequent tests. The
hypothesis is that batch testing may lead to variations in the number of
tests ordered, potentially influencing the overall healthcare
expenditure and efficiency.

\emph{72-hour Return}-- The rate of patients returning to the ED within
72 hours of their initial visit serves as a proxy for the quality of
care received. A high 72-hour return rate might indicate inadequate
treatment or diagnostic oversight during the initial visit. We measure
this outcome as the proportion of patients who have a subsequent ED
visit within 72 hours post-initial discharge.

\hypertarget{batching}{%
\subsubsection{Batching}\label{batching}}

In the context of our study, batching in diagnostic test ordering is
defined as placing multiple test orders for a single patient within a
5-minute interval. To ensure the robustness of our analysis, sensitivity
tests were conducted on this cutoff point, affirming that our results
are consistent and reliable irrespective of minor variations in the
batching definition.

Our study identifies and differentiates between two distinct strategies
of batching:

\begin{itemize}
\item
  Lab + Image Batch: This category of batch order includes instances
  where a patient's batched diagnostic tests comprise a combination of
  at least one imaging test (such as a contrasted CT scan,
  non-contrasted CT scan, X-ray, or ultrasound) and one laboratory test.
  This mix is particularly interesting as it suggests a comprehensive
  diagnostic approach, possibly catering to complex medical cases.
\item
  Image + Image Batch: This category encompasses batch orders where a
  patient's diagnostic tests consist of two or more distinct imaging
  tests. The reliance solely on imaging in these batches may indicate
  specific diagnostic pathways or a focus on conditions that primarily
  require visual diagnostic methods.
\end{itemize}

It is important to note the differing implications of these batching
strategies. While lab tests often can be processed concurrently with
imaging tests, causing minimal additional waiting time, a batch of
multiple imaging tests requires the patient's presence for each test,
potentially leading to significant increases in waiting time. Thus, the
composition of a batch order can provide valuable insights into the
physician's decision-making process, particularly in balancing between
patient waiting times and diagnostic uncertainty. This balance can be
especially crucial in the context of varying complexity levels of
patient encounters.

Figure 1 illustrates the occurrence rates of each batching type across
the top ten most frequently reported chief complaints, providing a
visual representation of how batching practices vary with different
patient presentations.

\begin{figure}[h]
  \centering
  \includegraphics[width=1\textwidth]{../outputs/figures/Lab and Image Batch Rates.png}
\begin{tablenotes}
\small
\item \textit{Figure 1 displays two statistics for the ten most common major chief complaint categories observed in our ED visit sample: the unadjusted batching rate for lab-image batching and image-image batching.}
\end{tablenotes}  
\end{figure}

\hypertarget{summary-statistics}{%
\subsection{Summary Statistics}\label{summary-statistics}}

\hyperref[tab:summary_statistics]{Table 1} presents a detailed overview
of the emergency department (ED) characteristics, patient demographics,
and medical tests for our baseline sample, derived from the data
collected during the study period. On average, the emergency department
manages a volume of approximately 24 patients (Mean = 24.18), indicating
a significant but manageable patient load. Key physiological markers
such as tachycardia (Mean = 0.19) and tachypnea (Mean = 0.09) are
prevalent among patients, albeit at varying degrees, highlighting
critical aspects of emergency care. Notably, a small proportion of
patients exhibit febrile (Mean = 0.02) and hypotensive (Mean = 0.01)
conditions, emphasizing the diversity of cases encountered in the ED
setting. The Emergency Severity Index (ESI) averages at 2.78, suggesting
that most cases fall within the moderate acuity level.

Common chief complaints in the emergency department paint a vivid
picture of patient needs and healthcare demands. The data reveals
abdominal complaints (Mean = 0.14), extremity complaints (Mean = 0.12),
and chest pain (Mean = 0.08) as the most frequently occurring, followed
by neurological (Mean = 0.08) and gastrointestinal issues (Mean = 0.08).
This information is crucial in understanding the primary reasons for ED
visits and the required resources for patient care.

The patient characteristics in our sample depict a diverse demographic
landscape. The average arrival age in the ED is approximately 58 years
(Mean = 58.33), indicating a predominantly middle-aged population. The
racial composition is predominantly White (Mean = 0.89), with smaller
proportions of Black (Mean = 0.04) and Asian (Mean = 0.03) patients.
Notably, a slight majority of the patients are female (Mean = 0.54),
offering insights into gender dynamics in healthcare utilization.

In terms of diagnostic tests, the data reveals a high reliance on lab
tests (Mean = 0.77), indicating their crucial role in patient diagnosis
and management. X-Ray (Mean = 0.47) and CT scans, both non-contrast
(Mean = 0.21) and contrast (Mean = 0.19), are also frequently employed,
underscoring the importance of imaging in modern emergency medicine.
Ultrasound usage (Mean = 0.11) and the practice of batch ordering tests
(Mean = 0.38) further reflect the operational aspects of the ED and the
strategies employed to manage patient care effectively.

\begin{table}[ht]
\centering
\caption{Summary Statistics}
\label{tab:summary_statistics}
\begin{tabular}{p{9cm}cccc}
\toprule
\textbf{} & \textbf{Mean} & \textbf{Q1} & \textbf{Median} & \textbf{Q3} \\
\midrule
\multicolumn{5}{l}{\textbf{ED Characteristics}} \\
ED Volume & 24.18 & 15 & 25 & 32 \\
Tachycardic & 0.19 & & & \\
Tachypneic & 0.09 & & & \\
Febrile & 0.02 & & & \\
Hypotensive & 0.01 & & & \\ 
ESI & 2.78 & 2 & 3 & 3 \\
Complaint: Abdominal & 0.14 & & & \\
Complaint: Extremity & 0.12 & & & \\
Complaint: Chest Pain & 0.08 & & & \\
Complaint: Neurological & 0.08 & & & \\
Complaint: Gastrointestinal & 0.08 & & & \\
\midrule
\multicolumn{5}{l}{\textbf{Patient Characteristics}} \\
Arrival Age & 58.33 & 44 & 62 & 75 \\
Race: White & 0.89 & & & \\
Race: Black & 0.04 & & & \\
Race: Asian & 0.03 & & & \\
Gender: Female & 0.54 & & & \\
\midrule
\multicolumn{5}{l}{\textbf{Tests}} \\
X-Ray & 0.47 & & & \\
Ultrasound & 0.11 & & & \\
Non-Contrast CT & 0.21 & & & \\
Contrast CT & 0.19 & & & \\
Lab & 0.77 & & & \\
Tests were Batch Ordered & 0.38 & & & \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small
\item \textit{Notes: This table reports summary statistics for the baseline sample of emergency department visits during the study period described in the text.}
\end{tablenotes}
\end{table}

Our empirical strategy closely follows the literature that relies on
quasi-random assignment of agents to cases, often referred to as the
``judges design.'' Papers in this literature typically exploit variation
in the sentencing leniency of judges who work in the same court.
Similarly, we explore batching variation across physicians who work in
the same emergency department. In its reduced form, under the assumption
of quasi-random assignment, this approach allows researchers to identify
the causal effect of being assigned to different types of physicians.
Under additional assumptions, an instrumental variable approach
identifies the causal effect of a given medical decision. We employ both
approaches and lay out their details in the next subsections.

@@

Contrary to most healthcare settings where patients exhibit choice, they
are predominantly passive in their physician assignment in the ED. In
most EDs, however, physicians have discretion in picking their patients.
In contrast, patients arriving at the Mayo Clinic ED are randomly
assigned to physicians via a rotational patient assignment algorithm
(Traub et al., 2016), which removes potential selection bias concerns
for our analyses. In essence, barring arrival time and shift-level
variation, the physician-to-patient matching can be deemed random.
\hyperref[tab:summary_statistics]{Table 2} displays that patient
encounters (regarding chief complaints and emergency severity) are
equitably distributed across physicians within our study's cohort.

\begin{table}[htbp]
    \centering
    \caption{Wald Test Results}
    \label{tab:wald_test}
    \begin{tabular}{p{10cm}cc}
        \toprule
        Chief Complaints & F-Statistic & $Pr(> F)$ \\
        \midrule
        Abdominal Complaints & 1.37 & 0.106 \\
        Back or Flank Pain & 1.00 & 0.451 \\
        Chest Pain & 0.98 & 0.476 \\
        Extremity Complaints & 0.97 & 0.495 \\
        Falls, Motor Vehicle Crashes, Assaults, and Trauma & 0.73 & 0.812 \\
        Gastrointestinal Issues & 0.98 & 0.480 \\
        Neurological Issue & 0.75 & 0.793 \\
        Shortness of Breath & 1.23 & 0.199 \\
        Skin Complaints & 1.05 & 0.388 \\
        Upper Respiratory Symptoms & 1.21 & 0.218 \\
        \midrule
        Emergency Severity & F-Statistic & $Pr(> F)$ \\
        \midrule
        ESI 1 or 2 & 1.09 & 0.346 \\
        ESI 3, 4, or 5 & 1.24 & 0.196 \\
        \bottomrule
    \end{tabular}
\begin{tablenotes}
\small
\item \textit{Table 2 reports the results of a Wald test which was conducted to assess the balance of chief complaints across providers in our dataset. A balanced distribution implies that complaints and severity are evenly distributed across providers, which we expect to be the case due to randomization. The Wald F-statistic and p-value are reported. Robust standard errors (type HC1) were used to account for potential heteroscedasticity in the data.}
\end{tablenotes}
\end{table}

\hypertarget{batch-tendency-construction}{%
\subsection{Batch Tendency
Construction}\label{batch-tendency-construction}}

@@

\begin{equation}
Batched_{i,t} = \alpha_0 + \alpha_{ym} + \alpha_{dt} + \alpha_{complaint} + \varepsilon_{i,t}
\end{equation}

Where \(Batched_{i,t}\) is a dummy variable equal to one if patient
\(i\) had their diagnostic tests batched on encounter that took place on
data \(t\). Fixed effects include year-month fixed effects,
\(\alpha_{ym}\), to control for time and seasonal variation in batching,
such as hospital-specific policies (e.g.~initiatives to eliminate excess
testing) or seasonality in ED visits. We also control for
``shift-level'' variations that include both physician scheduling and
patient arrival with day of week-time of day fixed effects,
\(\alpha_{dt}\). Chief complaint by severity fixed effects,
\(\alpha_{complaint}\), were also included to increase precision. As
stated earlier, these controls are what is required for our quasi-random
assignment assumption. Under the assumption that we have captured the
observables under which quasi-random assignment occurs in the ED, the
unexplained variation-- the physician's contribution-- resides in the
error term, \(\varepsilon_{i,t}\).

@@

\begin{equation}
Tendency_{i,j}^{phys} =
\frac{1}{N_{-i,j}} \sum_{i' \in \{J \backslash i\}}\hat{\varepsilon}_{i'}
\end{equation}

where \(\hat{\varepsilon}_{i'} = \hat{Batch}_{i'} - Batch_{i'}\) is the
residual from equation (1); \(J\) is the set of all ED encounters
treated by physician \(j\); and \(N_{-i,j} = |\{J \backslash i\}|\), the
number of cases that physician has seen that year, excluding patient
\(i\). This leave-out mean eliminates the mechanical bias that stems
from patient \(i\)'s own case entering into the instrument. The measure
is interpreted as the average (leave-out) batch rate of patient \(i\)'s
physician, relative to other physicians in that hospital-year-month,
hospital-day of week-time of day.

We document that the Mayo Clinic ED physicians exhibit wide, systematic
variation in their propensity to batch order diagnostic tests. Figure 2
graphs the histogram of batch-ordering frequency by physician for
popular chief complaints, highlighting that the variation in batching
differs systematically. \hyperref[tab:regression]{Table 3} presents the
``first stage'' in a regression table: being assigned to a 10 pp higher
tendency physician is associated with a 5 pp increase in the likelihood
of having tests batch-ordered in the ED. The F-statistic is 50 when all
controls and fixed effects are included. The coefficient is greater than
one because all emergency visits are used to construct the leniency
instrument, while the first stage is calculated using the baseline
sample only, which excludes the rare complaints.

\begin{table}[ht]
\centering
\caption{First Stage Regression Results}
\label{tab:regression}
\begin{tabular}{p{9cm}ccc}
\toprule
\textbf{} & \textbf{Model 1} & \textbf{Model 2} & \textbf{Model 3} \\
\midrule
Batching Tendency & 1.05 *** & 1.04 *** & 1.04 *** \\
 & (0.01) & (0.01) & (0.01) \\
\midrule
F statistic (full model) & 7.02 & 50.34 & 49.31  \\  
F (full model): p-value & $<0.001$ & $<0.001$ & $<0.001$ \\
\midrule
Num. obs. & 43327 & 43327 & 43327 \\
Seasonality and shift fixed effects? & Yes & Yes & Yes \\
Chief Complaint? & No & Yes & Yes \\
Patient observables? & No & No & Yes \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small
\item \textit{Estimates of the first stage for the baseline sample described in the text. Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects. Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter. Patient observables include sex dummy, race/ethnicity, and age bins. Column 3 corresponds to the baseline controls. Robust standard errors are clustered at the physician level.}
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}
\end{table}

@@

In this section we investigate whether the reduced-form effects observed
in Section 4.1 are due to differences in batch rates across providers or
due to other provider differences correlated with batch tendency We
start by studying reduced-form effects among patients with complaints
that are never prescribed opioids, as a ``placebo/falsification check.''
By way of example, consider a patient who arrives at the ED with a
urinary tact infection---a condition for which patients rarely undergo
imaging testing. For such patients, we should expect to see no impact of
batch tendency only if high batching and low batching physicians do not
systematically differ in other dimensions of care relevant to patient
outcomes. Conversely, if we do find a reduced-form effect for these
patients, then high batch tendency physicians must systematically differ
from low batch tendency physicians in other dimensions of care, beyond
batching.

To that end, we restrict attention to ED visits for complaints where
batching occurs no more than 10 percent of the time (recall that our
baseline sample only includes complaints with a \textgreater10 percent
batching rate). We estimate a reduced-form regression of each main
outcome on physician prescribing leniency for the subsample, following
equation (3). The results of this exercise are displayed in Table 4.
They show that in contrast to results for our main sample, the
association between physician tendency to batch and a given outcome is
statistically indistinguishable from zero and much smaller in magnitude
for the samples of patients who visit the ED with health conditions that
are rarely batched.

To further probe robustness, we perform the placebo exercise for a set
of non-opioid outcomes---in the form of homelessness, suicide, all-cause
mortality, and preventable hospitalizations---in online Appendix Table
G.7. Note that these outcomes are likely to be more sensitive to other
dimensions of physician care than our opioid-related outcomes because
the causal link to prescription opioid exposure is less tight, while
other margins of care may matter relatively more. Indeed, we do find a
sizable and statistically significant effect of being assigned to a
high-prescribing physician on preventable hospitalizations and all-cause
mor- tality for our placebo conditions. For patients with ED diagnoses
that are rarely prescribed an opioid (column 3), assignment to a
physician in the top decile of the opioid prescribing range increases
the likelihood of a preventable hospitalization within 3 years by 0.27
pp, or 1.8 percent.19

The failure of our placebo check for some non-opioid outcomes suggests
that there may be dimensions of care correlated with opioid prescribing
tendency that vary across ED physicians and may impact non-opioid
patient outcomes. While the set of potentially relevant dimensions of
care is very large and may include dimen- sions unobservable to the
researcher (such as, for example, instructions on medica- tion adherence
and use, physician-nurse scheduling that lead to complementarities), we
can make progress by accounting for the chief observable dimensions
likely to correlate highly with patient health outcomes in ED settings:
quality of care as mea- sured by immediate patient mortality, intensity
of procedures, and tendency to admit patients for inpatient
hospitalizations. In moving toward a well-identified IV design that
estimates the causal impact of an opioid prescription on patient
outcomes, we address these potential violations to the exclusion
restriction by estimating physician ``propensities'' along these
``non-focal'' (relative to opioid prescribing) dimensions and including
them as controls (details are provided in Section IIIE). This approach
follows recent advances from the judge stringency literature
(Mueller-Smith 2015; Bhuller et al.~2020), which deals with similar
concerns.

We recognize that given the multidimensionality of physician behavior,
potential for exclusion restriction violations remains even after
controlling for chief dimensions of behavior. For this reason, we also
present reduced-form results, and we keep a tight focus on
opioid-related outcomes, for which such concerns are less pronounced
relative to general health outcomes.
